---
title: "Project 2: Creating Vignettes for API's"
author: "Angelice Floyd"
date: "2023-10-08"
urlcolor: blue
---
Contens

 * Intro and Requirements   
 * Functions for API Interaction  
    * `PAData`
    * `PNUMData` 
    * `IAData`  
    * `EMPData`  
  * Data Exploration  
  * Wrap- Up  


This line of code creates the render function for R to be able to run wihout hitting the "Knit" button. 

```{r rndr, eval = FALSE, echo = FALSE}
rmarkdown::render("Project2.Rmd", 
              output_format = "github_document", 
              output_options = list(
                name_value_pairs = "value", 
                toc = TRUE,
                toc_depth = 4, 
                number_of_sections = TRUE, 
                df_print = "paged", 
                or_something = TRUE
                )
              )
```

 

#Introduction and Requirements 

## Introduction 

This project creates a vignette for reading and summarizing baseball statistics from a FEMA ( Federal Emergency Management Associate) API. The first section will consist of pulling in all of the necessary libraries needed to run the R program, contact the API from the MLB, create the exploratory data analysis from the data. The second sectino will create the functions for interacting with the API ( INSERT WICH PORTIONS OF THE OPTIONS THE USER CAN CHOOSE HERE ). The Third sectio will use the data from the API function interaction for some exploratory data analysis. Finally, there will be a wrap up to summarize what was done in the vignette.  

## Requirements 

This section loads the package required for interacting with the FEMA API, manipulating the data, and conducting exploratory data analysis.

```{r lib, message = FALSE, warning = FALSE}
library(tidyverse)
library(httr)
library(jsonlite)
```

#Functions for API Interactions

## Public Assistance Funded Projects Endpoint

The first function created will interact with the Public Assistance endpoint of the FEMA API. This data contiains numerical and categorical summary information on funcding projects that FEMA conducted geared towards public assistance. The next endpoint of this API contains the abbreviated form of the state name. Thus, for consistency, a two letter state abbreviation was added. 

```{r PAfun}
PAdata <- function(statename = "all") {
  #pull the public assistance summaries from the Public assistance endpoint
  
  outputAPI <- fromJSON(
    "https://www.fema.gov/api/open/v1/PublicAssistanceFundedProjectsSummaries" 
    )
  
   APIData <- outputAPI$PublicAssistanceFundedProjectsSummaries %>% 
     mutate(stateabb = state.abb[match(state,state.name)])
 
   #If the default state name of 'all is not included, then make sure that the 
   #chosen name is included, and filter, otherwise, print out an error
   if(statename != "all"){
     if(statename %in% APIData$state) {
       output <- APIData %>% filter(state == statename)
     }
     else{
       message <- print("Please try again and enter a valid state name")
       stop(message)
     }
   }
   else{ 
     output <- APIData
   }
  return(output)
}

PAData()
```

This next set of code modifies the endpoint to filter on pulling data from either the full state name, the  county name of the region, or the two letter abbreviated state name. Thus, the user could type in state name and the data filters on the state name value. However, if the user chooses to type in either the two letter state name or the county name, the API modifies to filter on the given value. Further, all of this can be typed in as one string under statename, where the default is all state values. Finally, the output pulls all of the relevant variables that will be used in the Exploratory Data analysis. 

```{r api2}
PNUMData <- function(statename = "all") {
  
  outputAPI <- fromJSON(
    "https://www.fema.gov/api/open/v1/PublicAssistanceFundedProjectsSummaries")
  APIData <- outputAPI$PublicAssistanceFundedProjectsSummaries %>% 
    mutate(stateabb = state.abb[match(state,state.name)])
  output <- APIData

  if(statename != "all"){
   
    if(statename %in% APIData$state) {
      
      output <- APIData %>% filter(state == statename)
    }
    
    else if( statename %in% APIData$county) {
      
      #Initial URL for the API 
      URL1 <- "https://www.fema.gov/api/open/v1/" 
      
      #Extended endpoint with the county filter
      endpoint <- "PublicAssistanceFundedProjectsSummaries?$filter=county%20eq"
      
      #final URL that will putll the above data 
      finURL <- paste0(URL1, endpoint,paste('%20%27',statename,'%27',sep=""),sep="")
      
      #Object that creates communicates to the FEMA API
      outputAPI <- fromJSON(finURL)
      
      #Finally, these objects create the dataset for the API Data
      output <- outputAPI$PublicAssistanceFundedProjectsSummaries
    }
    
   else if(length(statename) == 1) {
     short <- state.name[match(statename,state.abb)]
     
     #Initial URL for the API 
     URL1 <- "https://www.fema.gov/api/open/v1/" 
     
     #Extended endpoint with the county filter
     endpoint <- "PublicAssistanceFundedProjectsSummaries?$filter=state%20eq"
     
     #final URL that will putll the above data 
     finURL <- paste0(URL1, endpoint,paste('%20%27',short,'%27',sep=""),sep="")
     
     #Object that creates communicates to the FEMA API
     outputAPI <- fromJSON(finURL)
     
     #Finally, these objects create the dataset for the API Data
     output <- outputAPI$PublicAssistanceFundedProjectsSummaries
   }
   
  }
  else{
    
  }
  
  finoutput <- output %>% select(state,county,incidentType,county,educationApplicant,numberOfProjects,federalObligatedAmount)
 
 return(finoutput)
}

PNUMData()
```

Below is an output example for the county of Lamar in Alabama. 

```{r outexmple}
PNUMData("Lamar")
```

## Individual Assistance for Homeowners 

This next API Endpoint yeilds financial data from the FEMA Housing Assistance Program for Homeowners by zip code, county, city, and state for registered states. Since, I will not be excluding any data for the numerical analysis, all of the modifications and connections are doine in the same function.The first portion of the code is used to produce an initial connection to the  `HousingAssistanceOwners` API. This brings in the data for all of the states, counties, zip codes and cities- this is the default setting for the statename input. If the user chooses not to incorporate all of the regional based data, there are some options that they could type in which correspond to the following :  
  * Two letter state abbreviation: ex: AL  
  * specific full state name  
  * Zip code if the region is registered and in the dataset : This has to be in a character string.  
  * City: must be in all capitals. 

```{r IADTA}
IAData <- function(statename = "all") {
  
  outputAPI <- fromJSON(
  "https://www.fema.gov/api/open/v2/HousingAssistanceOwners" 
  )

  APIData <- outputAPI$HousingAssistanceOwners %>%
    rename("stateabb"=state) %>%
    mutate(state = state.name[match(stateabb,state.abb)])
  output <- APIData

  #If the default state name of 'all is not included, then make sure that the 
  #chosen name is included, and filter, otherwise, print out an error
  if(statename != "all"){
    if(statename %in% APIData$state) {
    output <- APIData %>% filter(state == statename)
    }
  
  else if( statename %in% APIData$zipCode) {
    
    #Initial URL for the API 
    URL1 <- "https://www.fema.gov/api/open/v2/" 
    
    #Extended endpoint with the zip code filter
    endpoint <- "HousingAssistanceOwners?$filter=zipCode%20eq"
    
    #final URL that will putll the above data 
    finURL <- paste0(URL1, endpoint,paste('%20%27',statename,'%27',sep=""),sep="")
    
    #Object that creates communicates to the FEMA API
    outputAPI <- fromJSON(finURL)
    
    #Finally, these objects create the dataset for the API Data
    output <- outputAPI$HousingAssistanceOwners
  }
  else if( statename %in% APIData$city) {
    
    #Initial URL for the API 
    URL1 <- "https://www.fema.gov/api/open/v2/" 
    
    #Extended endpoint with the zip code filter
    endpoint <- "HousingAssistanceOwners?$filter=city%20eq"
    
    #final URL that will putll the above data 
    finURL <- paste0(URL1, endpoint,paste('%20%27',statename,'%27',sep=""),sep="")
    
    #Object that creates communicates to the FEMA API
    outputAPI <- fromJSON(finURL)
    
    #Finally, these objects create the dataset for the API Data
    output <- outputAPI$HousingAssistanceOwners
  }
  
  else if(length(statename) == 1) {
    short <- state.name[match(statename,state.abb)]
    
    #Initial URL for the API 
    URL1 <- "https://www.fema.gov/api/open/v1/" 
    
    #Extended endpoint with the county filter
    endpoint <- "PublicAssistanceFundedProjectsSummaries?$filter=state%20eq"
    
    #final URL that will putll the above data 
    finURL <- paste0(URL1, endpoint,paste('%20%27',short,'%27',sep=""),sep="")
    
    #Object that creates communicates to the FEMA API
    outputAPI <- fromJSON(finURL)
    
    #Finally, these objects create the dataset for the API Data
    output <- outputAPI$PublicAssistanceFundedProjectsSummaries
  }
  
else{
  
}
  } 
  return(output)
}
```

Below is an output example of the data for a zip code in Arkansas

```{r output}
IAData('99776')
```

## Emergency Management Performance Grant endpoint

The final interaction with the FEMA API will be connecting to the Emergency Management Performance Grant endpoint.This produces a dataset that shows the amount of grant funding provided to regional emergency management agencies in order to assist them with  their emergency preparedness initiatives over a specific length of time.  The modifications to the API will be the state, and the beginning date of the project funded, as it will be interesting to be able to filter the data pertaining to the times where different states conducted their emergency preparedness projects. 

First, this set of code creates a list of possible combinations of state names and start dates for reference. 

```{r indx}

outputAPI <- fromJSON(
  "https://www.fema.gov/api/open/v2/EmergencyManagementPerformanceGrants" 
)

APIDataIN <- outputAPI$EmergencyManagementPerformanceGrants %>%
  mutate(stateabb = state.abb[match(state,state.name)]) %>%  separate(projectStartDate,c("startdate","starttime"),sep = "T",convert=TRUE, remove = FALSE) %>% separate(projectEndDate,c("enddate","endtime"), sep = "T", convert = TRUE, remove = FALSE) %>% unite("index",state,startdate, remove = FALSE)


print(tibble(index = unique(APIDataIN$index)) %>% separate(index,c("state","startdate"),sep="_") %>% mutate(stateabb = state.abb[match(state,state.name)]))
```

Next, this set of code creates a function that pulls the Emergency Management Performance Grant data from the endpoint. There are some options for the user.  
  * Enter  nothing -`EMPData()`- and all of the data ( unfiltered ) is pulled from the endpoint.  
  * Enter the state name ( full or abbreviated), but not the start date - `EMPData(statename = "AL")`, then if the full name is give, the filtered data from the first bullet is output, but if the abbreviated version is given, the full name of the abbreviated version is found and the endpoint is modified to pull from the converted abbreviation and the data fror the corresponding state with all possible dates is produced.   
  * Eter the start date but not the state name `EMPData(stdate = "2014-10-01")` ,the endpoint, the startdate is first parsed from the projectStartDate variable ( since this variable includes both the startdate and a starttime), united with the starttime, and then endpoint is modified to find the re-combined value within the projectStartDate.  
  * If both the state name and start date are entered `EMPData(stdate = "2014-10-01",statename = "AL")`, the program follows the logic of the previous two bullets, and the endpoint is modified to serach for the two values.  
  
```{r EMPFunction}
EMPData <- function(statename = "all",stdate = "all") {
  
  outputAPI <- fromJSON(
    "https://www.fema.gov/api/open/v2/EmergencyManagementPerformanceGrants" 
  )
  
  APIDataIN <- outputAPI$EmergencyManagementPerformanceGrants %>%
    mutate(stateabb = state.abb[match(state,state.name)]) %>%  separate(projectStartDate,c("startdate","starttime"),sep = "T",convert=TRUE, remove = FALSE) %>% separate(projectEndDate,c("enddate","endtime"), sep = "T", convert = TRUE, remove = FALSE)
  
  output <- APIDataIN
  
  #If the default state name of 'all is not included, then make sure that the 
  #chosen name is included, and filter, otherwise, print out an error
  if(statename != "all" & stdate == "all"){
    if(statename %in% APIDataIN$state) {
      output <- APIData %>% filter(state == statename)
    }
    
    else if(length(statename) == 1) {
      short <- state.name[match(statename,state.abb)]
      
      #Initial URL for the API 
      URL1 <- "https://www.fema.gov/api/open/v2/" 
      
      #Extended endpoint with the county filter
      endpoint <- "EmergencyManagementPerformanceGrants?$filter=state%20eq"
      
      #final URL that will putll the above data 
      finURL <- paste0(URL1, endpoint,paste('%20%27',short,'%27',sep=""),sep="")
      
      #Object that creates communicates to the FEMA API
      outputAPI <- fromJSON(finURL)
      
      #Finally, these objects create the dataset for the API Data
      output <- outputAPI$EmergencyManagementPerformanceGrants %>%  separate(projectStartDate,c("startdate","starttime"),sep = "T",convert=TRUE, remove = FALSE) %>% separate(projectEndDate,c("enddate","endtime"), sep = "T", convert = TRUE, remove = FALSE)
    }
  } 
  else if(stdate != "all" & statename == "all") {
    dateval <- paste0(stdate,"T",APIDataIN$starttime[1])
    
    #Initial URL for the API 
    URL1 <- "https://www.fema.gov/api/open/v2/" 
    
    #Extended endpoint with the county filter
    endpoint <- "EmergencyManagementPerformanceGrants?$filter=projectStartDate%20eq"
    
    #final URL that will putll the above data 
    finURL <- paste0(URL1, endpoint,paste('%20%27',dateval,'%27', sep=""),sep="")
    
    #Object that creates communicates to the FEMA API
    outputAPI <- fromJSON(finURL)
    
    #Finally, these objects create the dataset for the API Data
    output <- outputAPI$EmergencyManagementPerformanceGrants %>%  separate(projectStartDate,c("startdate","starttime"),sep = "T",convert=TRUE, remove = FALSE) %>% separate(projectEndDate,c("enddate","endtime"), sep = "T", convert = TRUE, remove = FALSE)
    
  }
    else if(stdate != "all" & statename != "all"){
      if(statename %in% APIDataIN$state){
        dateval <- paste0(stdate,"T",APIData$starttime[1])
      
      #Initial URL for the API 
      URL1 <- "https://www.fema.gov/api/open/v2/" 
      
      #Extended endpoint with the county filter
      endpoint <- "EmergencyManagementPerformanceGrants?$filter=projectStartDate%20eq"
      
      #final URL that will putll the above data 
      finURL <- paste0(URL1, endpoint,paste('%20%27',dateval,'%27%20and%20state%20eq%20%27',statename,"%27",sep=""),sep="")
      
      
      #Object that creates communicates to the FEMA API
      outputAPI <- fromJSON(finURL)
      
      #Finally, these objects create the dataset for the API Data
      output <- outputAPI$EmergencyManagementPerformanceGrants %>%  separate(projectStartDate,c("startdate","starttime"),sep = "T",convert=TRUE, remove = FALSE) %>% separate(projectEndDate,c("enddate","endtime"), sep = "T", convert = TRUE, remove = FALSE)
      
    }
  else if(length(statename)==1){
    #Take the start date given value and convert it into the Project start Date format
    
    dateval <- paste0(stdate,"T",APIDataIN$starttime[1])
    long <- state.name[match(statename,state.abb)]
    
    #Initial URL for the API 
    URL1 <- "https://www.fema.gov/api/open/v2/" 
    
    #Extended endpoint with the county filter
    endpoint <- "EmergencyManagementPerformanceGrants?$filter=projectStartDate%20eq"
    
    #final URL that will putll the above data 
    finURL <- paste0(URL1, endpoint,paste('%20%27',dateval,'%27%20and%20state%20eq%20%27',long,"%27",sep=""),sep="")
    
    #Object that creates communicates to the FEMA API
    outputAPI <- fromJSON(finURL)
    
    #Finally, these objects create the dataset for the API Data
    output <- outputAPI$EmergencyManagementPerformanceGrants
  }
}
  return(output)
}

```

Below are examples of the options available to the user for this function. 

```{r options, eval = FALSE}
EMPData()
EMPData(statename = "AK")
EMPData(statename = "Alaska")
EMPData(stdate = "2014-07-01")
EMPData(stdate = "2014-07-01",statename = "AK")
EMPData(stdate = "2014-07-01",statename = "Alaska")
```

## Data Exploration

For the Data Exploration, There a few things that I am interested in exploring. For the 
